Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-05 E-Hypergraphs-LICS24.tex
intro

\begin{abstract}
The technique of equipping graphs with an equivalence relation, called equality saturation, has recently proved both powerful and practical in program optimisation, particularly for SMT solvers. 
We give a categorical semantics for \emph{e-graphs} over algebraic theories, a generalisation of the structures which are the subject of this technique, and to their rewriting. 
The proposed framework consists of Cartesian categories enriched over a semilattice,  further generalised to symmetric monoidal categories, to account for monoidal theories.
The aim is to permit the optimisation of a larger class of programs, based on string diagrams.  
Furthermore, we present a sound and complete combinatorial representation of morphisms in such a category, based on a generalisation of hypergraphs,  which we call \emph{e-hypergraphs}, which have the usual advantage that many structural equations are absorbed into a general notion of isomorphism. 
\end{abstract}

@@

\subsection{Semantics via Semilattice Enriched Categories}

@@

Recalling the correspondence between DAG representations of algebraic terms and  string-diagrams for morphisms of a suitable Cartesian category
makes perspicuous the correspondence between (acyclic) e-graphs and morphisms of a suitably enriched Cartesian category, taken as string diagrams~\cite{noauthor_09083347_nodate,joyal_geometry_1991, mellies_functorial_2006}. In Figure~\ref{fig:egraph-strings} we extend the usual vocabulary of string diagrams for symmetric monoidal categories with the additional generator  
\[
\infer{\phi_1 + \ldots + \phi_n: A \to B}{\{\phi_i: A \to B\}_{i \in \{1,\ldots,n\}}}
\]
The ability to take formal (non-empty) joins of morphisms is used to model the equivalence class structure of e-graphs.

@@

Note how the natural level of generality for our string diagrams is not Cartesian, but monoidal: the generators of a Cartesian category are required in order to embed e-graphs, but are orthogonal to the addition of formal joins. 
Thus we generalise e-graphs from algebraic to monoidal theories, giving rise to a  host of new potential applications, which we outline in the conclusion of this paper.   
\\ DAN: THIS IS A BIT UNCLEAR. WHY IS THIS THE NATURAL LEVEL THE GENERALITY? IT'S NOT QUITE OBVIOUS.

@@

In order to \textit{implement} generalised e-graphs, rewriting must be performed on concrete representations of the corresponding string diagrams.  String diagrams can be considered equivalently as either topological objects (\textit{i.e.}, taken modulo ``connectivity'') or as a 2-dimensional syntax quotiented by the equations of an SMC. For example, we have the following equivalences between diagrams. 
%\[(f_1 \otimes f_2) ; (g_1 \otimes g_2) = (f_1 ; g_1) \otimes (f_2 ; g_2)\] 
\[
	\scalebox{0.6}{
	\tikzfig{categorical-semantics/interchange}
	}
\]
This representation makes string diagrams unamenable to efficient implementation due to the difficulty of calculating the quotient. 
A large body of research work shows how this issue can be solved by taking a different approach, representing string diagrams as hypergraphs~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.  Here, the generators $c_i$ become vertices, connected via hyper-edges.  Thus the expected quotient becomes simply hypergraph isomorphism. With appropriate restrictions on the form these hypergraphs can take, this approach can be used to \textit{characterise} the free symmetric monoidal category generated by some $c_i$.

@@

\subsection{Soundness and Completeness}
The main technical result of the paper is a proof soundness and completeness of the combinatorial representation for semilattice enriched SMCs, extending the results of \cite{bonchi_string_2022-2} for plain SMCs. While the structural equalities of SMCs are factored out in the representation,  important structural equalities arising from the enrichment (see the equations of Figure \ref{fig:string-equations}) are not, and should not be. They represent the un/sharing of subdiagrams with respect to the e-box structure,  which is precisely what allows for the compact representation of equivalence classes.  Instead, we consider DPO-rewrites implementing both the structural equalities for enrichment (which involve the e-box structure) and the equalities arising from the generating monoidal theory (which do not).

@@

Our use of string diagrams is also central to the intuition behind our work.  The string diagrams we work with are left essentially as informal reasoning devices, however there exists a breadth of work on similar diagrammatic formalisms for similar classes of categories.  We mention here work on hierarchical string diagrams \cite{ghica_hierarchical_2023},  functorial boxes \cite{mellies_functorial_2006},  proof nets for compact closed categories with biproducts \cite{duncan_generalised_2009}, and recent work on tape diagrams for rig categories with biproducts \cite{bonchi_tape_nodate}.

@@

\section{Categorical Semantics of E-Graphs}
CHRIS: INTRO TO SECTION: CLARIFY THAT WE GIVE E-GRAPHS AS CARTESIAN CASE,  BUT ALSO GENERALIZE TO MONOIDAL CASE. 
CHRIS: CITE HASEGAWA - TRACE MONOIDAL CATEGORIES, AND PERDIX.

>>>>

\begin{abstract}
The technique of equipping graphs with an equivalence relation, called equality saturation, has recently proved both powerful and practical in program optimisation, particularly for SMT solvers. 
We give a categorical semantics to these structures, called \emph{e-graphs}, in terms of Cartesian categories enriched over a semilattice. 
We show how this semantics can be generalised to monoidal categories, which opens the door to new applications of e-graph techniques, from algebraic to monoidal theories.
Finally, we present a sound and complete combinatorial representation of morphisms in such a category, based on a generalisation of hypergraphs (\emph{e-hypergraphs}).
They have the usual advantage that many of their structural equations are absorbed into a general notion of isomorphism. 
\end{abstract}

@@

\subsection{Semilattice Enriched Categories}

@@

Recalling the correspondence between DAG representations of algebraic terms and  string-diagrams for morphisms of a suitable Cartesian category
makes perspicuous the correspondence between (acyclic) e-graphs and morphisms of a suitably enriched Cartesian category, taken as string diagrams~\cite{noauthor_09083347_nodate,joyal_geometry_1991, mellies_functorial_2006}. We extend the usual vocabulary of string diagrams for symmetric monoidal categories with the additional generator  
\[
\infer{\phi_1 + \ldots + \phi_n: A \to B}{\{\phi_i: A \to B\}_{i \in \{1,\ldots,n\}}}
\]
The ability to take formal (non-empty) joins of morphisms is used to model the equivalence class structure of e-graphs.

@@

Note that the motivating example of our work, e-graphs, requires the Cartesian structure. 
However, we will see how all the core of the theory of e-hypergraph rewriting only requires the monoidal structure. 
Thus we can readily generalise e-graphs from algebraic to monoidal theories, giving rise to a  host of new potential applications, which we briefly outline in the conclusion of this paper.

@@

In order to \textit{implement} generalised e-graphs, rewriting must be performed on concrete representations of the corresponding string diagrams.  String diagrams can be considered equivalently as either topological objects (\textit{i.e.}, taken modulo ``connectivity'') or as a 2-dimensional syntax quotiented by the equations of an SMC. For example, we have the following equivalences between diagrams. 
%\[(f_1 \otimes f_2) ; (g_1 \otimes g_2) = (f_1 ; g_1) \otimes (f_2 ; g_2)\] 
\[
	\scalebox{0.6}{
	\tikzfig{categorical-semantics/interchange}
	}
\]
This representation makes string diagrams unamenable to efficient implementation due to the difficulty of calculating the quotient. 
A large body of research work shows how this issue can be solved by taking a different approach, representing string diagrams as hypergraphs~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.  Here, the generators $c_i$ become vertices, connected via hyper-edges.  Thus the expected quotient becomes simply hypergraph isomorphism. With appropriate restrictions on the form these hypergraphs can take, this approach can be used to \textit{characterise} the free symmetric monoidal category generated by some~$c_i$.

@@

\subsection{Soundness and Completeness}
The main technical result of the paper is a proof of soundness and completeness of the combinatorial representation for semilattice enriched SMCs, extending the results of \cite{bonchi_string_2022-2} for plain SMCs. While the structural equalities of SMCs are factored out in the representation,  important structural equalities arising from the enrichment (see the equations of Figure \ref{fig:string-equations}) are not, and should not be. They represent the un/sharing of subdiagrams with respect to the e-box structure,  which is precisely what allows for the compact representation of equivalence classes.  Instead, we consider DPO-rewrites implementing both the structural equalities for enrichment (which involve the e-box structure) and the equalities arising from the generating monoidal theory (which do not).

@@

Our use of string diagrams is also central to the intuition behind our work.  
DAN: INFORMAL REASONING DEVICES DOESN'T SOUND GOOD. WHAT DO YOU MEAN? 
The string diagrams we work with are left essentially as informal reasoning devices, however there exists a breadth of work on similar diagrammatic formalisms for similar classes of categories.  We mention here work on hierarchical string diagrams \cite{ghica_hierarchical_2023},  functorial boxes \cite{mellies_functorial_2006},  proof nets for compact closed categories with biproducts \cite{duncan_generalised_2009}, and recent work on tape diagrams for rig categories with biproducts \cite{bonchi_tape_nodate}.

@@

\section{Categorical Semantics of E-Graphs}
CHRIS: INTRO TO SECTION: CLARIFY THAT WE GIVE E-GRAPHS AS CARTESIAN CASE,  BUT ALSO GENERALIZE TO MONOIDAL CASE. 
CHRIS: CITE HASEGAWA - TRACE MONOIDAL CATEGORIES, AND PERDIX. 
DAN: I SUGGEST TO CITE HASEGAWA IN FURTHER WORK, FOR GENERALISING THE THEORY TO TRACED MC.
